Modeling Intervention Strategies for United States TB Control

### 2013-10-14 in_progress/presentations/paper/draft.tex
added capability to change LTBI Cure Cost when generating tables for intervention cost
added figure captions

\begin{figure}
\includegraphics[scale=0.25]{figures/HillModelFlowChart.png}
\end{figure}

@@

\section{Methods}
\subsection{Basic Structure}
Both the Hill Model (Basic Hill Model) and an extended version of the Hill Model (Extended Hill 
Model) were implemented in 
\texttt{R} as a system of differential equations, which were solved via the
$\texttt{lsoda}$ routine. The systems of differential equations usedto capture basic disease dynamics in the Basic Hill Model and the Extended Hill Model are
shown below in Figure 2 and Figure 3 respectively. In the Extended Hill Model,
more differential equations were implemented to add additional tracking and
economic modeling. These equations are detailed in the Appendix, Section 7.2.

\begin{figure}
\includegraphics[scale=0.75]{figures/BasicHillEquations.pdf}
\end{figure}

@@

\subsection{Additional Tracking Capabilities}
In order to refine the tracking capabilities of the
Hill Model, the original differential equations used to describe TB spread were
separated into their component parts and each section was tracked separately.
These components were made into compartments, tracked by differential equations
detailed in the appendix. These equations allowed the model to track the total
number of TB cases, TB deaths, and  natural deaths. Further, the model tracks,
the sourced cost on the US HCS. Progressions into
active TB due to activations of LTBI or exogenous infection were tracked,
allowing for sourced incidence data to be generated. The entering cases of LTBI
(acute or chronic) were also tracked. In the case of intervention testing, the
number of cured and untreated cases of entering LTBI were both tracked. Cured
cases of LTBI entering, TB deaths, total TB cases, and total cost were also
tracked discounted at 3\% annually. This discounting was converted to a
continuous differential equation for use within the model. In general, incidence
data is calculated in the same way in the extended hill as it was in the basic
Hill.

@@

Further implementations of the \texttt{C++} model were made that deviated from
the basic Hill Model. Specifically, the acute latent compartment in the basic
Hill Model is a vestigal necessity of compartmental modeling and not reflective
of the biology of tuberculosis, and an agent based model was implemented that
more closely respects the biology of tuberculosis. Results from this model or
the strict Hill model did not differ significantly. 
\section{Results}
\subsection{Basic Population Breakdown}
The additional tracking capabilities offered several key insights into US TB
dynamics. Figure~\ref{fig:incPlotSourced} shows the yearly incidence of US TB, broken
down into infection source. It can be seen that the majority of the US TB load
is driven by activations of FB LTBI, followed by USB LTBI activations. This data
further agrees with the conclusions drawn by Hill, Becerra, and Castro about the
necessity of LTBI treatment in any valuable intervention strategy. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.5]{incPlotSourced}
  \end{center}
  \caption{Sourced yearly incidence data generated by the extended Hill Model.}
  \label{fig:incPlotSourced}
\end{figure}
Further, figure~\ref{fig:costPlotSourced} shows a similarly sourced plot, but analyzing the
final US HCS costs due to TB. One can see that in this plot, roughly half of the
US TB HCS costs are due to activations of LTBI. In this data, it is appropriate
to think of the costs due to acute LTBI are due to exogenous re-inection,
whereas active TB costs due to chronic LTBI are activations of LTBI into active
TB. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.35]{costPlotSourced}
  \end{center}
  \caption{Sourced US HCS economic TB load. Note that this data only illustrates
    the load due to treating active TB, but illustrates where the infections
    driving this cost come from.}
  \label{fig:costPlotSourced}
\end{figure}
Note that both of these plots underestimate the impact LTBI activations play in
the spread of TB, as every LTBI activation to infectious TB contributes not only
to incidence and costs directly, but also indirectly by causing additional
future cases, which is not captured in these graphs. Further, there is also US
HCS costs due to treating LTBI, which is not illustrated in these graphs. 
\subsubsection{Basic Reproduction Number}
The basic reproduction number of FB or USB cases of infectious TB was also
estimated by this system. Using the best-fit values of parameters from the 
Hill model, we found the average number of secondary infections arising from
a single case of infectious TB to be 0.423 and 0.370 for USB and FB populations
respectively.  These values differ because in the model, there are different
estimates for death rates and the rate of TB treatment and self-cure in the USB and FB populations, 
so the average infectious period differs between the two populations.
Extrapolating these results, we can think of the
total number of secondary infections over 100 years due to a FB or USB
infectious TB case as describing a geometric series in a large population.
Presuming there are no overlaps in infectious contacts, if a single case of
infectious TB in either population infects $p_f, p_u$ new cases,
respectively, then we can say that over the course of 100 years, the total
number of cases infected will follow a geometric series. This analysis predicts
that over 100 years, one USB infection will lead to $1.04$ subsequent infections,
whereas one FB infection will lead to $.83$ subsequent infections. Experimentally, 
these data were also analyzed, with results of $1.03$ and $.64$, respectively. 
Full calculations are included in the appendix.

@@

Several other intervention strategies analyzed by the Hill Model were refined
with the Extended Hill, and economic properties about each of them were tracked.
The results for these interventions, which are less effective than curing
entering LTBI cases across the board, are given in the appendix. 
\subsection{Agent Based Evaluation}
The Agent Based Model allowed the statistical properties of the system to be
analyzed and verified. In particular, they illustrated that the deterministic
Hill model provides a robust and consistent statistical measure of TB epidemic
behavior in the US conditions. We found that the distribution of incidence and
final population sizes were normal, with mean accurate to the deterministic
model and standard deviations given in table X, in the appendix.

\subsection{Sensitivity Analysis}
To account for uncertainty of input parameter values and to gain insight about the
most influential parameters in the model, we used Latin Hypercube Sampling to 
analyze the deterministic model, implemented with the 'lhs' package in R.  We varied 
the 16 parameters from the original Hill model according to triangular distributions centered
around their best-fit values, and two additional treatment cost parameters according to uniform distributions.  We obtained similar results to the Hill model 
analyzing non-economic outcomes, as expected for validation.  For economic 
outcomes, we found the parameter $f$, the fraction of FB arrivals with LTBI,
to be highly correlated with the projected overall cost burden of TB in the United States
over the next 100 years.  Another highly influential parameter in the model was $\sigma^{L}$, 
the treatment rate for chronic LTBI.  However, while increasing $\sigma^{L}$ significantly
decreases the cost burden due to Active TB treatment, it increases the cost burden due to Latent TB treatment by a greater amount, given the model estimates that Latent TB treament and Active
TB treatment health care costs are approximately \$700 and \$14,000 per case cured respectively.  
A full description of the Latin Hypercube Sampling analysis and tables of Partial Rank Correlation Coefficient (PRCC) results for all parameters are given in the Appendix, section 7.6.  \\

@@

% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Mon Oct 14 00:36:15 2013
\begin{table}
\centering
\begin{tabular}{|r|cccc|}
  \hline
 & red5 & red10 & red25 & red50 \\ 
  \hline
2000 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  2025 & 1115.12 & 2231.70 & 5592.43 & 11227.57 \\ 
  2050 & 3406.41 & 6821.27 & 17117.18 & 34447.71 \\ 
  2075 & 4976.77 & 9967.15 & 25021.33 & 50389.58 \\ 
  2100 & 5942.02 & 11900.90 & 29880.03 & 60189.49 \\ 
   \hline
\end{tabular}
\caption{Cases of TB Averted by Reducing Incoming LTBI by X percent} 
\label{tab:caRed}
\end{table}

% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Mon Oct 14 00:36:15 2013
\begin{table}
\centering
\begin{tabular}{|r|cccc|}
  \hline
 & red5 & red10 & red25 & red50 \\ 
  \hline
2000 & ND & ND & ND & ND \\ 
  2025 & 67908.81 & 67855.46 & 67654.07 & 67332.53 \\ 
  2050 & 28938.23 & 28878.70 & 28699.15 & 28404.13 \\ 
  2075 & 21044.88 & 20989.16 & 20820.53 & 20541.77 \\ 
  2100 & 18130.11 & 18076.11 & 17912.95 & 17642.64 \\ 
   \hline
\end{tabular}
\caption{Cost Per Case Averted by Reducing Incoming LTBI by X percent (in dollar per case)} 
\label{tab:cpcaRed}
\end{table}

>>>>

\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=0.25]{figures/HillModelFlowChart.png}
  \end{center}
  \caption{Schematic of the Hill Model.}
  \label{fig:hillModelSchematic}
\end{figure}

@@

\section{Methods}
\subsection{Basic Structure}
Both the Hill Model (Basic Hill Model) and an extended version of the Hill Model (Extended Hill 
Model) were implemented in 
\texttt{R} as a system of differential equations, which were solved via the
$\texttt{lsoda}$ routine. The systems of differential equations used to capture basic disease dynamics in the Basic Hill Model are shown in Figure 2. In the Extended Hill Model,
more differential equations were implemented to add additional tracking and
economic modeling. These equations are detailed in the Appendix, Section 7.2.

\begin{figure}
  \begin{center}
    \includegraphics[scale=0.75]{figures/BasicHillEquations.pdf}
  \end{center}
  \caption{System of Differential Equations given in the Hill Model.}
  \label{fig:hillEquations}
\end{figure}

@@

\subsection{Additional Tracking Capabilities}
In order to refine the tracking capabilities of the
Hill Model, the original differential equations used to describe TB spread were
separated into their component parts and each section was tracked separately.
These components were made into compartments, tracked by differential equations
detailed in the appendix. These equations allowed the model to track the total
number of TB cases, TB deaths, and  natural deaths. Further, the model tracks,
the sourced cost on the US HCS. Progressions into
active TB due to activations of LTBI or exogenous infection were tracked,
allowing for sourced incidence data to be generated. The entering cases of LTBI
(acute or chronic) were also tracked. In the case of intervention testing, the
number of cured and untreated cases of entering LTBI were both tracked. Cured
cases of LTBI entering, TB deaths, total TB cases, and total cost were also
tracked discounted at 3\% annually. This discounting was converted to a
continuous differential equation for use within the model. In general, incidence
data is calculated in the same way in the Extended Hill Model as it was in the Basic
Hill Model.

@@

Further implementations of the \texttt{C++} model were made that deviated from
the basic Hill Model. Specifically, the acute latent compartment in the basic
Hill Model is a vestigal necessity of compartmental modeling and not reflective
of the biology of tuberculosis, and an agent based model was implemented that
more closely respects the biology of tuberculosis. Results from this model or
the strict Hill model did not differ significantly. 
\section{Results}
\subsection{Basic Population Breakdown}
The additional tracking capabilities offered several key insights into US TB
dynamics. Figure~\ref{fig:incPlotSourced} shows the yearly incidence of US TB, broken
down into infection source. It can be seen that the majority of the US TB load
is driven by activations of FB LTBI, followed by USB LTBI activations. This data
further agrees with the conclusions drawn by Hill, Becerra, and Castro about the
necessity of LTBI treatment in any valuable intervention strategy. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.5]{incPlotSourced}
  \end{center}
  \caption{Sourced yearly incidence data generated by the extended Hill Model.}
  \label{fig:incPlotSourced}
\end{figure}
Further, figure~\ref{fig:costPlotSourced} shows a similarly sourced plot, but analyzing the
final US HCS costs due to TB. One can see that in this plot, roughly half of the
US TB HCS costs are due to activations of LTBI. In this data, it is appropriate
to think of the costs due to acute LTBI are due to exogenous re-inection,
whereas active TB costs due to chronic LTBI are activations of LTBI into active
TB. 
\begin{figure}[h]
  \begin{center}
    \includegraphics[scale=.35]{costPlotSourced}
  \end{center}
  \caption{Sourced US HCS economic TB load. Note that this data only illustrates
    the load due to treating active TB, but illustrates where the infections
    driving this cost come from.}
  \label{fig:costPlotSourced}
\end{figure}
Note that both of these plots underestimate the impact LTBI activations play in
the spread of TB, as every LTBI activation to infectious TB contributes not only
to incidence and costs directly, but also indirectly by causing additional
future cases, which is not captured in these graphs. Further, there is also US
HCS costs due to treating LTBI, which is not illustrated in these graphs. 
\subsubsection{Basic Reproduction Number}
The basic reproduction number of FB or USB cases of infectious TB was also
estimated by this system. Using the best-fit values of parameters from the 
Hill model, we found the average number of secondary infections arising from
a single case of infectious TB to be 0.423 and 0.370 for USB and FB populations
respectively.  These values differ because in the model, there are different
estimates for death rates and the rate of TB treatment and self-cure in the USB and FB populations, 
so the average infectious period differs between the two populations.
Extrapolating these results, we can think of the
total number of secondary infections over 100 years due to a FB or USB
infectious TB case as describing a geometric series in a large population.
Presuming there are no overlaps in infectious contacts, if a single case of
infectious TB in either population infects $p_f, p_u$ new cases,
respectively, then we can say that over the course of 100 years, the total
number of cases infected will follow a geometric series. This analysis predicts
that over 100 years, one USB infection will lead to $1.04$ subsequent infections,
whereas one FB infection will lead to $.83$ subsequent infections. Experimentally, 
these data were also analyzed, with results of $1.03$ and $.64$, respectively. 
Full calculations are included in the Appendix.

@@

Several other intervention strategies analyzed by the Hill Model were refined
with the Extended Hill, and economic properties about each of them were tracked.
The results for these interventions, which are less effective than curing
entering LTBI cases across the board, are given in the appendix. 
\subsection{Agent Based Evaluation}
The Agent Based Model allowed the statistical properties of the system to be
analyzed and verified. In particular, they illustrated that the deterministic
Hill model provides a robust and consistent statistical measure of TB epidemic
behavior in the US conditions. We found that the distribution of incidence and
final population sizes were normal, with mean accurate to the deterministic
model and standard deviations given in table X, in the Appendix.

\subsection{Sensitivity Analysis}
To account for uncertainty of input parameter values and to gain insight about the
most influential parameters in the model, we used Latin Hypercube Sampling to analyze the deterministic model, implemented with the \texttt{lhs} package in \texttt{R}.  We varied 
the 16 parameters from the original Hill model according to triangular distributions centered
around their best-fit values, and two additional treatment cost parameters according to uniform distributions.  We obtained similar results to the Hill model 
analyzing non-economic outcomes, as expected for validation.  For economic 
outcomes, we found the parameter $f$, the fraction of FB arrivals with LTBI,
to be highly correlated with the projected overall cost burden of TB in the United States
over the next 100 years.  Another highly influential parameter in the model was $\sigma^{L}$, 
the treatment rate for chronic LTBI.  However, while increasing $\sigma^{L}$ significantly
decreases the cost burden due to Active TB treatment, it increases the cost burden due to Latent TB treatment by a greater amount, given the model estimates that Latent TB treament and Active
TB treatment health care costs are approximately \$700 and \$14,000 per case cured respectively.  
A full description of the Latin Hypercube Sampling analysis and tables of Partial Rank Correlation Coefficient (PRCC) results for all parameters are given in the Appendix, Section 7.6.  \\

@@

% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Mon Oct 14 00:36:15 2013
\begin{table}
\centering
\begin{tabular}{|r|cccc|}
  \hline
 & red5 & red10 & red25 & red50 \\ 
  \hline
2000 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
  2025 & 1115.12 & 2231.70 & 5592.43 & 11227.57 \\ 
  2050 & 3406.41 & 6821.27 & 17117.18 & 34447.71 \\ 
  2075 & 4976.77 & 9967.15 & 25021.33 & 50389.58 \\ 
  2100 & 5942.02 & 11900.90 & 29880.03 & 60189.49 \\ 
   \hline
\end{tabular}
\caption{Cases of TB Averted by Reducing Incoming LTBI by X percent (Cure=\$800)} 
\label{tab:caRed}
\end{table}

% latex table generated in R 3.0.1 by xtable 1.7-1 package
% Mon Oct 14 00:36:15 2013
\begin{table}
\centering
\begin{tabular}{|r|cccc|}
  \hline
 & red5 & red10 & red25 & red50 \\ 
  \hline
2000 & ND & ND & ND & ND \\ 
  2025 & 67908.81 & 67855.46 & 67654.07 & 67332.53 \\ 
  2050 & 28938.23 & 28878.70 & 28699.15 & 28404.13 \\ 
  2075 & 21044.88 & 20989.16 & 20820.53 & 20541.77 \\ 
  2100 & 18130.11 & 18076.11 & 17912.95 & 17642.64 \\ 
   \hline
\end{tabular}
\caption{Cost Per Case Averted by Reducing Incoming LTBI by X percent (in dollar per case, Cure=\$800)} 
\label{tab:cpcaRed}
\end{table}
